\chapter{Appendix 2: R Code Used in this Project}

#EM Estimation of Bivariate Poisson HMM Parameters

This code is adapted from that found in @zucchini to support bivariate Poisson models

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk('../code/hmmBivPois.R')
```

```{r bi.pois.HMM.EM, eval=FALSE}
```

\newpage

#Calculation of Forward and Backward Probabilities

This code is adapted from that found in @zucchini to support bivariate Poisson models

```{r bi.pois.HMM.lalphabeta, eval=FALSE}
```

\newpage

#Monte-Carlo Simulation of Volume Buckets

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk('../code/VPIN-Sim.R')
```

```{r vpin.sim, eval=FALSE}
```

\newpage

#Data Loading and Cleansing of NYSE TAQ Data

Following instructions found in the highfrequency package manual

http://highfrequency.herokuapp.com

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk('../code/TAQ-Load-SPY.R')
```

```{r taq.load, eval=FALSE}
```

\newpage

#Technique to calculate VPIN from Time Bars

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk('../code/VPIN-Timebars.R')
```

```{r vpin.timebars, eval=FALSE}
```

\newpage

#Jump method to Identify Number of Clusters using K-Means

Thanks to Gareth M James for making this code available

http://www-bcf.usc.edu/~gareth/research/jump

```{r cache=FALSE, echo=FALSE}
knitr::read_chunk('../code/jump.R')
```

```{r jump, eval=FALSE}
```

