---
title: "TAQ Data Loading and Cleansing"
output: html_document
---

See following links for more info:

http://people.stern.nyu.edu/jhasbrou/Teaching/PrinciplesSecTrade/TAQ/taqExercise1.html

http://highfrequency.herokuapp.com/index.html

```{r echo=FALSE, message = FALSE, warning=FALSE}
library(zoo)
library(xts)
library(highfrequency)
library(Defaults)
library(TTR)
library(quantmod)
library(timeDate)

conversionRequired = FALSE;
setwd("/Users/nick/Documents/RStudioProjects/Dissertation/code")
datasource = "/Users/nick/Documents/RStudioProjects/Dissertation/data/raw";
datadestination = "/Users/nick/Documents/RStudioProjects/Dissertation/data/xts";
datadestination_cleaned = "/Users/nick/Documents/RStudioProjects/Dissertation/data/xts_cleaned";
```

```{r}
from = "2014-03-04";
to = "2014-03-04";

tickers = c("ACW", "ALG","HZO","ITG")

tradeColNames <- c("DATE","TIME_M","EX","SYMBOL","SYM_SUFFIX","COND","SIZE","PRICE","TR_STOPIND","CORR","TR_SEQNUM","TR_SOURCE","TR_RF")

quoteColNames <- c("DATE","TIME_M","EX","SYMBOL","SYM_SUFFIX","BID","BIDSIZ","OFR","OFRSIZ","QU_COND","BIDEX","ASKEX","QU_SEQNUM","NATBBO_IND","NASDBBO_IND","QU_CANCEL","QU_SOURCE")

```

###Data Loading
```{r warning=FALSE}
if(conversionRequired){
  #TODO: check the source of the G127 NA errors
  suppressWarnings(
    convert(from, to, datasource, datadestination, trades=TRUE,
          quotes=TRUE,ticker=tickers, dir=FALSE, extension="csv",
          header=TRUE,tradecolnames=tradeColNames,quotecolnames=quoteColNames,
          format="%Y%m%d %H:%M:%OS")
  )
}

options("digits.secs"=3); #Shows milliseconds
tdata = TAQLoad(tickers="HZO", from=from, to=to, trades=T, quotes=F, 
                       datasource=datadestination)

qdata = TAQLoad(tickers="HZO", from=from, to=to, trades=F, quotes=T, 
                       datasource=datadestination)

head(tdata)

head(qdata)
```

###Data Cleanup and Outlier Removal
```{r}
#TODO: investigate R timezone handling (low priority)

tdata = tradesCleanup(tdataraw = tdata,exchanges="D", report = FALSE)
qdata = quotesCleanup(qdataraw = qdata,exchanges="T", report = FALSE, maxi=25) #be a bit more aggressive in cleaning up quotes with abnormally large spreads (default is 50)

tdataAfterFinalCleanup = tradesCleanupFinal(qdata = qdata, tdata = tdata)
```

###Matching Trades and Quotes
```{r}
tqdata = matchTradesQuotes(tdataAfterFinalCleanup,qdata);

plot.xts(tqdata[, "PRICE"], type = "bars")
lines(tqdata[, "OFR"], col="red")
lines(tqdata[, "BID"], col="green")
```


###Liquidity Measures

- Inferred trade direction

A vector which has values 1 or (-1) if the inferred trade direction is buy or sell respectively.

```{r}
tradeDirections = getTradeDirection(tqdata)
plot(tradeDirections)
```

- Effective Spread

$\mbox{effective spread}_t = 2*D_t*(\mbox{PRICE}_{t} - \frac{(\mbox{BID}_{t}+\mbox{OFR}_{t})}{2}),$

```{r}
effectiveSpread = tqLiquidity(tqdata,type="es")
plot(effectiveSpread)
```

- Realised Spread

$\mbox{realized spread}_t = 2*D_t*(\mbox{PRICE}_{t} - \frac{(\mbox{BID}_{t+300}+\mbox{OFR}_{t+300})}{2})$

where $D_t$ is 1 (-1) if $trade_t$ was buy (sell) (see Boehmer (2005), Bessembinder (2003)). Note that the time indication of $\mbox{BID}$ and $\mbox{OFR}$ refers to the registered time of the quote in seconds.

```{r}
realisedSpread = tqLiquidity(tqdata,tdata,qdata,type="rs")
plot(realisedSpread)
```

- Price Impact

$\mbox{price impact}_t = \frac{\mbox{effective spread}_t - \mbox{realized spread}_t}{2}$

```{r}
price_impact = tqLiquidity(tqdata,tdata,qdata,type="price_impact")
plot(price_impact)
```
